Optimistic Atomic Multicast in One Communication Delay

### 2011-08-03 trunk/textos/optpaxos/draft_dropbox.tex

In Section \ref{sec:baseline}, we presented a straight-forward version of our atomic multicast protocol, which, despite being easily understandable, is open to several optimizations, which are described hereafter.  In Section \ref{sec:nullondemand}, we present a more latency-efficient approach to ensure that processes will eventually receive all barriers necessary to deliver each message; in Section \ref{sec:paxos}, we give more details regarding how consensus is achieved within only three communication steps; in Section \ref{sec:parallel}, we show that different instances of consensus may be voted upon and decided in parallel, further reducing the time needed for consensus and, in Section \ref{sec:optdel}, we present our optimistic delivery algorithm, which may deliver messages in one single communication step. Furthermore, in Section \ref{sec:discard}, we show how to relax the uniform validity property in order to reduce the final delivery time and the number of mistakes made by the optimistic delivery.

\subsection{Requesting empty messages}
\label{sec:nullondemand}

@@

@@

Even if a reliable multicast primitive is used to send every message to each one of its destinations, guaranteeing its arrival, messages from different senders may arrive in different orders at different destinations, so consensus is necessary to decide which one should be considered by all processes. However, we can predict the final delivery order using the timestamps assigned to the messages by their senders: if each process $p$ waits long enough before proposing some message $m$, every message $m' : m'.ts < m.ts$ will arrive eventually and $p$ will be able to propose them in the same order of their initial timestamps, achieving total order without needing any consensus for that.

The problem is then how to define the length of such wait window for each message $m$ such that it guarantees that every message prior to $m$ will have already been received when the window time has elapsed. As the message delay is unpredictable in asynchronous systems, we make an optimistic assumption:

@@

More formally, if $p \in G$, we can define $w(p)$ as $max_{p' \in sendersTo(G)}(\delta(p,p')+\epsilon(p,p'))$, where $\delta(p,p')$ is the maximum time a message takes to go all the way from $p$ to $p'$ and $\epsilon(p,p')$ is the difference between the clocks of $p$ and $p'$, so that clock deviations can be accounted for\footnote{If this difference is less than zero, it means that the clock value in $p'$ is higher than that in $p$, so $p$ actually has to wait less time for messages from $p'$, as they will have higher timestamps because of the clock deviation.}. The clock deviation is mentioned in this assumption because the timestamp $m.ts$ of each message $m$ is assigned by its sender according to its local clock and we want that, once a message $m$ has been proposed, no message $m' : m'.ts < m.ts$ arrives afterwards. If our optimistic assumption considered only the message transmission delay, such property would not be guaranteed by it.

However, the assumption may not hold, so deliveries made based on it have to be confirmed. There would be then two deliveries for each message: an optimistic one, done within one communication step, and a conservative one, which may follow one of the algorithms described earlier. The difference is that, after a message $m$ was \rmdel{}ed at $p$, instead of proposing it, $p$ inserts it into an $optPending$ set, which is sorted in ascending order of the timestamps of the messages in it. Then, $p$ waits until its own wallclock has a value greater than $m.ts.rtc + w(p)$, after which $p$ both proposes $m$ and \opt{}s (optimistically delivers) it. If \emph{A2} holds, then all messages sent by $m.src$ and received after $m$ has been \opt{}ed will have a timestamp greater than $m.ts$.

The optimistic assumption allows for even further improvement. If it holds, the timestamp of the messages will never be changed (i.e. l. \ref{algline:chgts} of Alg. \ref{algorithm:nullondemand} will never be executed). Therefore, if Algorithm \ref{algorithm:nullondemand} is being used, there is no need to wait until a message $m$ has been decided to know its final timestamp and only then request barriers to groups in $blockers(m)$. After the wait window for $m$ has elapsed, such barrier request can already be sent.

If \emph{A2} fails to hold and some message $m$ is received by a process $p$ after a message \mbox{$m' : m'.ts > m.ts$} has been already \opt{}ed and proposed by $p$, then the optimistic delivery algorithm made a mistake, which the application can figure out by the different delivery orders and take action to correct whatever problem this mistake might have caused. Besides, when such thing happens, it means that the timestamp of $m$ has been changed by the conservative delivery algorithm and that, maybe, a new barrier request must be sent to $blockers(m)$ when this new timestamp is defined.

The delivery time of each message $m$ will now depend on whether the assumption \emph{A2} holds or not while $m$ is being handled by the algorithm. If it does hold, it will be \opt{}ed in the correct order within $w(p)$ and, as a barrier request was done in parallel with the conservative delivery algorithm, it will take $w(p) + 2\tconsm + \delta$ to \cons{} $m$ in the worst case. If it does not hold, after $w(p)$, $m$ may be \opt{}ed in an invalid order and, since its timestamp may have changed, a new barrier request may have to be done. Therefore, the worst case conservative message delivery time when the optimistic assumption does not hold and some message is delivered out of order would be $2w(p) + 4\tconsm + \delta$.

Figure \ref{fig:optdel} illustrates both cases, when the optimistic assumption holds and otherwise. Three groups are shown: $G$, $G'$ and $G_b$, where $sendersTo(G) = \{G_b\}$, $sendersTo(G') = \{G\}$ and \mbox{$sendersTo(G_b) = \emptyset$}. A message $m : m.src = G \wedge m.dst = \{G, G'\}$ is \amcast{} and, at the same time, a barrier request is sent to the only group in $blockers(m)$, which is $G_b$. In (a), the optimistic assumption holds, so no message $m' : m'.ts < m.ts$ arrives after the proposal of $m$; in (b), however, it is necessary to make a second barrier request.

\begin{figure} 
  \centering
  \includegraphics[width=0.8\linewidth]{images/optdel}
  \caption{Execution examples when the optimistic assumption holds (a) and when it fails to hold (b)}
  \label{fig:optdel}
\end{figure}

@@

We can further optimize the delivery algorithm. One major problem with proposing messages with consensus is that the previous algorithms wait until a consensus instance has been finished to start a new one, so that no message is proposed twice. However, even if a message is proposed and accepted twice, each process may choose to consider only the first time (i.e. the consensus instance with lowest id) when a message $m$ has been proposed. This would require some more local processing and could cause some unnecessary traffic due to messages being proposed in different consensus instances. The worst case delivery time of a message $m$, when using the optimistic delivery and barrier requests, would be $2w(p) + 2\tconsm + \delta$. It must be noted, however, that this represents the case when the optimistic assumption fails to hold and $m$ has its timestamp changed. Otherwise, the conservative delivery time would be $w(p) + \tconsm + \delta$.

The basic idea would be that, whenever a process $p$ has a pending message $m$ that it received, it will propose it in some instance $I$ as soon as its clock has a value greater than $m.ts.rtc + w(p)$. As $p$ cannot foresee whether $m$ will be decided in $I$ or not, it keeps the double $(m,I.id)$ in a $trying$ set. When $I$ terminates, $(m,I.id)$ is removed from $trying$ and $p$ checks whether $m$ has been decided -- which might have happened also in some instance $I' \neq I$ in the meantime -- by checking its $\decided$ set. If not, $p$ proposes $m$ again in some consensus instance $I''$ and inserts $(m,I''.id)$ into $trying$.

Even if the consensus instances are run in parallel, we consider that they make a callback to \emph{decide$_g$()} in ascending order of instance id. This can be easily done by the consensus implementation using a sequence number. Whenever a message $m$ is agreed upon in some consensus instance $I$, any later decision of $m$ in some instance \mbox{$I' : I'.id > I.id$} is merely ignored.

Although ensuring termination of consensus in an asynchronous system is not possible \cite{fischer1985idc}, the Paxos \cite{lamport1998ptp} algorithm can guarantee the termination of an instance $I$ as long as some assumptions are held:

\begin{itemize}
  \item the maximum message delay bound $\delta$ is known;
  \item at some point in the execution of $I$, there is a leader which does not fail until a value has been chosen\footnote{Lamport demonstrates in \cite{lamport1998ptp} that, if the time needed to elect a leader is $T_{el}$, the time needed to conclude a consensus instance would be at most $T_{el} + 9\delta$ after the last leader failure during the execution of $I$.};
  \item during the whole execution of $I$, more than half of the processes are correct, that is, if the number of faulty processes is $f$ and the total number of processes is $n$, $f < \lceil \frac{n}{2} \rceil$;
  \item enough messages are successfully received, that is, for each phase of the execution of $I$, a majority of processes succesfully receive the leader's message or the leader successfully receives the reply (be it a confirmation or a negative acknowledgement) from a majority of processes.
\end{itemize}

Paxos uses the abstraction of \emph{proposers}, \emph{acceptors} and \emph{learners}. In short, the acceptors are the processes which have to agree upon some value given by the proposers. The learners are those who are notified about which value has been accepted. With a leader, Paxos can achieve consensus in most instances, except the first one since the last leader change, in $3\delta$ -- forwarding a proposal to the leader, sending it to the other processes (acceptors) and receiving a confirmation. Each acceptor can send the confirmation to every learner, so that each learner can figure out by itself that a value has been agreed upon once it receives a confirmation from a majority of acceptors.

In the context of our \amcast{} primitive, the first communication step of Paxos -- forwarding a message $m$ to the leader -- is already done by means of \rmcast{}ing each message to a set of processes that includes every process in $m$'s group of origin. The last phase, which consists of every learner receiving the confirmation of a majority of acceptors, can also include the $\delta$ contained in the conservative message delivery time we analysed before: instead of waiting for a message to be decided and only then notifying other groups, the acceptors of a group can also send the confirmation to processes in other groups, so that also they can infer the first group's decision right away. This implies that, for each consensus instance run within a group $G$, each process in the groups of $receiversFrom(G)$ would be a learner for that instance. Figure \ref{fig:paxosmanylearnergroups} illustrates this: process $p_1$ from group $G$, whose leader is $p_3$, \amcast{}s $m$; in the last phase of the consensus protocol, the processes in $G' \in receiversFrom(G)$ receive the confirmation message from the acceptors of $G$, so no extra communication step is needed to notify them about the final timestamp of $m$.

As $m$ was \rmcast{} in the beginning, there is no need to send all of its contents again in the consensus instance. Instead, only the timestamp of $m$ is proposed, since this is the information that must be agreed upon -- and which may change if the optimistic assumption does not hold and some other message of higher timestamp is delivered before $m$. In any case, as all messages agreed upon in $G$ are notified to $G'$, also the processes of $G'$ are able to infer the final timestamp of each message from $G$.

As we are making an optimistic assumption -- that each $p$ waits for a time $w(p)$ long enough so that no timestamps must be changed -- for the message delivery, the barriers are being requested to other groups in parallel, since $m$ is also \rmcast{} to $blockers(m)$ in the first communication step of our protocol. If such assumption holds, the barriers received will have a greater timestamp than $m$ and, thus, the conservative delivery of $m$ will take place $w(p) + 2\delta$ after it has been sent. If the clocks are perfectly synchronized and $w(p)$ does correspond to the maximum interprocess communication delay $\delta$, the time needed to conservatively deliver a message using this algorithm would be $3\delta$.

Finally, we can guarantee a conservative delivery time of $3\delta$, if no message had its timestamp changed, because the barrier requests for that message were made based on its initial timestamp. If the timestamp has changed, then a new barrier request may have to be made to ensure the delivery of that message. Unfortunately, to guarantee the delivery of the message, such request can be done only after the final timestamp has been decided: once the source group of a message $m$ noticed a change in $m.ts$, it sends another barrier request to $blockers(m)$. This request is handled in a way very similar to a conservative delivery, in the sense that, once it is received by a process, a null message is proposed, decided and then sent back to the destinations of $m$. This would take extra $3\delta$, so the worst case conservative delivery time would be $6\delta$.

Note, however, that for this worst case scenario to happen it would be necessary for the leader $p$ of group $G$ to: (1) not know some value greater than or equal to $w(p)$; (2) have received and proposed some message $m' : m'.ts > m.ts$ before $m$ and (3) not have received any barrier with value greater than the new timestamp of $m$. Even after changing the timestamp of $m$, it may be not necessary to request a new barrier from each group in $sendersTo(G)$, because some other barrier -- possibly requested because of some other message -- with a value higher than the new timestamp of $m$ may have been received already.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/paxos3d}
  \caption{Deciding the final timestamp and notifying other groups in three communication steps}
  \label{fig:paxosmanylearnergroups}
\end{figure}

@@

Although unlikely to happen, the worst case delivery time of \amcast{} when using Paxos with a leader for consensus and requesting barriers in the beginning of the protocol is $6\delta$. It is so because some messages may need to have their initial timestamp changed as a possible consequence of the optimistic assumption \emph{A2} not holding. However, a simple way to eliminate the possibility of having to change timestamps is by not considering messages whose delivery order does not follow the order of the initial timestamps.

Although discarding messages which do not follow the initial timestamp order conflicts with the properties we have defined for the \amcast{} primitive, we describe it here because then every message would be delivered within $3\delta$, in the worst case. The validity property would now be changed to:

\textbf{Optimistic Uniform Validity}: if a process \amcast{}s $m$ and the optimistic assumption \emph{A2} holds, then one of the correct processes that is a destination of $m$ eventually \cons{}s $m$.

The problem with changing the original uniform validity property is that it could make the multicast primitive impracticable for some applications. Nevertheless, some other applications not only do not require messages to be delivered if they arrive out of order, but also it is better not to deliver such messages at all -- such as real-time streams of audio or video or online real-time multiplayer games.

>>>>

In Section \ref{sec:baseline}, we presented a straight-forward version of our atomic multicast protocol, which, despite being easily understandable, is open to several optimizations, which are described hereafter.  In Section \ref{sec:nullondemand}, we present a more latency-efficient approach to ensure that processes will eventually receive all barriers necessary to deliver each message; in Section \ref{sec:parallel}, we show that different instances of consensus may be voted upon and decided in parallel, further reducing the time needed for consensus; in Section \ref{sec:paxos}, we give more details regarding how consensus is achieved within only three communication steps and, in Section \ref{sec:optdel}, we present our optimistic delivery algorithm, which may deliver messages in one single communication step. Furthermore, in Section \ref{sec:compositets}, we present a better way to define the timestamps of the messages and, in Section \ref{sec:discard}, we show how to relax the uniform validity property in order to reduce the final delivery time and the number of mistakes made by the optimistic delivery.

\subsection{Requesting barriers}
\label{sec:nullondemand}

@@

We can further optimize the delivery algorithm. One major problem with Algorithm \ref{algorithm:deliveryminimal} is that the processes wait until a consensus instance has been finished to start a new one, so that no message is proposed twice. However, even if a message is proposed and accepted twice, each process may choose to consider only the first time (i.e. the consensus instance with lowest id) when a message $m$ has been proposed. This would require some more processing and could cause some unnecessary traffic due to messages being proposed in different consensus instances, but it would reduce the average time needed to achieve consensus. %The worst case delivery time of a message $m$, when using the optimistic delivery and barrier requests, would be $2w(p) + 2\tconsm + \delta$. It must be noted, however, that this represents the case when the optimistic assumption fails to hold and $m$ has its timestamp changed. Otherwise, the conservative delivery time would be $w(p) + \tconsm + \delta$.

The basic idea would be that, whenever a process $p$ from group $g$ has an undecided message $m$ in $\localmsgs$, $p$ will propose $m$ in some instance $k$ as soon as $m$ is \rmdel{}ed. As $p$ cannot foresee whether $m$ will be decided in $k$ or not, it keeps the double $(m,k)$ in a $trying$ set. When $k$ terminates, $(m,k)$ is removed from $trying$ and $p$ checks whether $m$ has been decided -- which might have happened also in some instance $k' \neq k$ in the meantime -- by checking its $\decided$ set. If not, $p$ proposes $m$ again in some consensus instance $k''$ and inserts $(m,k'')$ into $trying$. If a message $m$ is agreed upon in some consensus instance $k$, the decision of $m$ in any instance \mbox{$k' : k' > k$} is ignored.

As processes do not block while running a consensus instance, there is no guarantee that the consensus instances will terminate in the correct order, which is: when a process $p$ from group $g$ decides $(k, val)$, $p$ has already decided $(k',val')$ for any instance $k' < k$, unless $k$ was the first consensus instance within $g$. However, we assume that this order is followed, which can be provided by the consensus implementaion by simply using a sequence number. This way, we can abstract the details of decision ordering.

Although ensuring termination of consensus in an asynchronous system is not possible \cite{fischer1985idc}, the Paxos algorithm~\cite{lamport1998ptp} can guarantee the termination of a consensus instance $k$ as long as some assumptions are held:

\begin{itemize}
  \item the maximum message delay bound $\delta$ is known;
  \item at some point in the execution of $k$, there is a leader which does not fail until a value has been chosen\footnote{Lamport demonstrates in \cite{lamport1998ptp} that, if the time needed to elect a leader is $T_{el}$, the time needed to conclude a consensus instance would be at most $T_{el} + 9\delta$ after the last leader failure during the execution of $k$.};
  \item during the whole execution of $k$, more than half of the processes are correct, that is, if the number of faulty processes is $f$ and the total number of processes is $n$, then $f < \lceil \frac{n}{2} \rceil$;
  \item enough messages are successfully received, that is, for each phase of the execution of $k$, a majority of processes succesfully receive the leader's message or the leader successfully receives the reply (be it a confirmation or a negative acknowledgement) from a majority of processes.
\end{itemize}

Paxos uses the abstraction of \emph{proposers}, \emph{acceptors} and \emph{learners}. In short, the acceptors are the processes which have to agree upon some value given by the proposers. The learners are those who are notified about which value has been accepted. With a leader, Paxos can achieve consensus in most instances, except the first one since the last leader change, in $3\delta$ -- the tree phases are: a proposal is sent to the leader, which forwards it to the other processes (acceptors) and receives a confirmation. Each acceptor can send the confirmation to every learner, so that each learner can figure out by itself that a value has been agreed upon once it receives a confirmation from a majority of acceptors.

In the context of our \amcast{} primitive, we consider that every correct process of a group is proposer, acceptor and learner in every consensus instance run by that group. Besides, we assume that one of such processes has been elected as leader. The first phase of Paxos -- forwarding a proposal to the leader -- is already done when the source process $p$ of a message $m$ \rmcast{}s it to all process in $group(p)$. The last phase message can also be sent to all destination processes of $m$, so that they can infer that $m$ has been accepted in its source group. For the consensus instance deciding each message $m$ within a given group, each process in $m.dst$ would also be a learner for that instance. This would be done instead of the processes of $group(p)$ waiting for $m$ to be decided to only then \rmcast{} it of to all its destinations. By doing this, every message $m$ could be decided and sent to every destination process within 3$\delta$ from when $m$ was created.

One problem could be that, if timestamps are defined as in Algorithm \ref{algorithm:deliveryminimal}, each destination process would need to know the contents of $\decided$ to infer the final timestamp of any message $m$ sent by group $g$. However, we are assuming that there is a leader process, by which all proposals have to pass. Such leader can, therefore, predict what will be contained in $\decided$ and set the final timestamp of $m$ before sending it to the acceptors in $g$ in the second phase, so that the confirmation message from the acceptors in the last phase of Paxos would already contain the final timestamp of $m$. Besides, the leader could request the barriers necessary for the delivery of $m$ at the beginning of the second phase of Paxos.

Figure \ref{fig:paxosmanylearnergroups} illustrates the execution: process $p_1$ from group $g$, whose leader is $p_3$, \amcast{}s $m$. In the first phase of the algorithm, $m$ is \rmcast{} to $p_3$. Then, in the second phase, $p_3$ has to change $m.ts$ because some other message with higher timestamp has passed by $p_3$ previously. Then, still in the second phase, $p_3$ sends $m$, with its new timestamp, to all acceptors, i.e., all processes of $g$. Finally, in the last phase, the processes in $h$ receive the confirmation message from the acceptors of $g$, already containing the final timestamp of $m$.

%As $m$ was \rmcast{} in the beginning, there is no need to send all of its contents again in the consensus instance. Instead, only the timestamp of $m$ is proposed, since this is the information that must be agreed upon -- and which may change if the optimistic assumption does not hold and some other message of higher timestamp is delivered before $m$. In any case, as all messages agreed upon in $G$ are notified to $G'$, also the processes of $G'$ are able to infer the final timestamp of each message from $G$.

%As we are making an optimistic assumption -- that each $p$ waits for a time $w(p)$ long enough so that no timestamps must be changed -- for the message delivery, the barriers are being requested to other groups in parallel, since $m$ is also \rmcast{} to $blockers(m)$ in the first communication step of our protocol. If such assumption holds, the barriers received will have a greater timestamp than $m$ and, thus, the conservative delivery of $m$ will take place $w(p) + 2\delta$ after it has been sent. If the clocks are perfectly synchronized and $w(p)$ does correspond to the maximum interprocess communication delay $\delta$, the time needed to conservatively deliver a message using this algorithm would be $3\delta$.

%Finally, we can guarantee a conservative delivery time of $3\delta$, if no message had its timestamp changed, because the barrier requests for that message were made based on its initial timestamp. If the timestamp has changed, then a new barrier request may have to be made to ensure the delivery of that message. Unfortunately, to guarantee the delivery of the message, such request can be done only after the final timestamp has been decided: once the source group of a message $m$ noticed a change in $m.ts$, it sends another barrier request to $blockers(m)$. This request is handled in a way very similar to a conservative delivery, in the sense that, once it is received by a process, a null message is proposed, decided and then sent back to the destinations of $m$. This would take extra $3\delta$, so the worst case conservative delivery time would be $6\delta$.

%Note, however, that for this worst case scenario to happen it would be necessary for the leader $p$ of group $G$ to: (1) not know some value greater than or equal to $w(p)$; (2) have received and proposed some message $m' : m'.ts > m.ts$ before $m$ and (3) not have received any barrier with value greater than the new timestamp of $m$. Even after changing the timestamp of $m$, it may be not necessary to request a new barrier from each group in $sendersTo(G)$, because some other barrier -- possibly requested because of some other message -- with a value higher than the new timestamp of $m$ may have been received already.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/paxos3d}
  \caption{Deciding the final timestamp and notifying all destination groups in three communication steps}
  \label{fig:paxosmanylearnergroups}
\end{figure}

@@

Even if a fifo reliable multicast primitive is used to send every message to each one of its destinations, guaranteeing its arrival in fifo order, messages from different senders may arrive in different orders at different destinations, so consensus is necessary to decide which order should be considered by all processes. However, we can predict the final delivery order using the initial timestamps assigned to the messages by their senders: if each process $p$ waits long enough before proposing some message $m$, every message $m' : m'.ts < m.ts$ will arrive eventually and $p$ will be able to propose them in the same order of their initial timestamps, achieving total order without needing any consensus for that.

The problem is then how to define the length of such wait window for each message $m$ such that every message prior to $m$ will have already been received when the window time has elapsed. As the message delay is unpredictable in asynchronous systems, we make an optimistic assumption:

@@

More formally, for $p$ in group $g$, we can define $w(p)$ as $max_{q \in sendersTo(g)}(\delta(p,q)+\epsilon(p,q))$, where $q$ is a process, $\delta(p,q)$ is the maximum time a message takes to go all the way from $p$ to $q$ and $\epsilon(p,q)$ is the difference between the clocks of $p$ and $q$\footnote{If this difference is less than zero, it means that the clock value in $q$ is higher than that in $p$, so $p$ actually has to wait less for messages from $q$, as they will have higher timestamps because of the clock deviation.}. The clock deviation is mentioned in this assumption because the timestamp of each message $m$ is assigned according to the local clock of its sender and we want that, once $m$ has been proposed, no message $m' : m'.ts < m.ts$ arrives afterwards. If our optimistic assumption considered only the message transmission delay, such property would not be guaranteed. However, the assumption may not hold, so deliveries made based on it have to be confirmed. There would be then two deliveries for each message: an optimistic one, done within one communication step, and a conservative one, based on consensus and barriers.

The optimistic delivery works as follows: after a new message $m$ has been \rmdel{}ed at $p$, instead of proposing it, $p$ %inserts it into an $optPending$ set. Then, $p$ 
waits until its own wallclock has a value greater than $m.ts + w(p)$. At this moment, if \emph{A2} holds, then all messages that could possibly have an initial timestamp lower than $m$ have already been received by $p$. Therefore, such messages can be \optdel{}ed in the order of their initial timestamps.

If every process $q$ of $g$ has estimated $w(q)$ correctly, then all processes of $g$ will propose all messages from such group in the same order, which is that of the initial timestamps. If this happens, there will be no timestamp changes and the optimistic delivery will be correct. Not changing timestamps allows for further improvement: if barriers are being requested for each message $m$, as described in Section \ref{sec:nullondemand}, the barrier requests may be sent to $blockers(m)$ at the same time as $m$ is \rmcast{} by its sender.

If \emph{A2} fails to hold and some message $m$ is received by some process $p$ after a message $m'$ has been already \optdel{}ed by $p$, where $m.ts < m'.ts$, then the optimistic delivery algorithm has made a \textit{mistake}. Such mistake can be figured out by the application from the difference between the optimistic delivery order and the conservative one. The application should then correct whatever problems this mistake might have caused. Besides, when such a mistake happens, it means that the timestamp of a message may have changed and that a new barrier request may have to be sent to $blockers(m)$ once the new timestamp has been defined.

%The delivery time of each message $m$ will now depend on whether the assumption \emph{A2} holds or not while $m$ is being handled by the algorithm. If it does hold, it will be \optdel{}ed in the correct order within $w(p)$ and, as a barrier request was done in parallel with the conservative delivery algorithm, it will take $w(p) + 2\tconsm + \delta$ to \consdel{} $m$ in the worst case. If it does not hold, after $w(p)$, $m$ may be \optdel{}ed in an invalid order and, since its timestamp may have changed, a new barrier request may have to be done. Therefore, the worst case conservative message delivery time when the optimistic assumption does not hold and some message is delivered out of order would be $2w(p) + 4\tconsm + \delta$.

%Figure \ref{fig:optdel} illustrates both cases, when the optimistic assumption holds and otherwise. Three groups are shown: $G$, $G'$ and $G_b$, where $sendersTo(G) = \{G_b\}$, $sendersTo(G') = \{G\}$ and \mbox{$sendersTo(G_b) = \emptyset$}. A message $m : m.src = G \wedge m.dst = \{G, G'\}$ is \amcast{} and, at the same time, a barrier request is sent to the only group in $blockers(m)$, which is $G_b$. In (a), the optimistic assumption holds, so no message $m' : m'.ts < m.ts$ arrives after the proposal of $m$; in (b), however, it is necessary to make a second barrier request.

% \begin{figure} 
%   \centering
%   \includegraphics[width=0.8\linewidth]{images/optdel}
%   \caption{Execution examples when the optimistic assumption holds (a) and when it fails to hold (b)}
%   \label{fig:optdel}
% \end{figure}

@@

\subsection{Timestamp composition}
\label{sec:compositets}

\textit{\textbf{I'm not sure of how good this `optimization' is; maybe it's not worth putting on the paper...}}

Barrier requests for each message $m$ may be sent before $m$'s final timestamp is decided. However, if $m.ts$ changes, new barrier requests will probably be deemed necessary. For instance, let $m$ and $m'$ be two messages sent by different processes of the same group $g$, where their initial timestamps are such that $m.ts < m'.ts$. Assume also that barrier requests have been sent at the same time as such messages were being sent. Suppose that some process $p$ from $g$, due to some asynchronous behavior of the system, received $m'$ first, proposed it and had $m'$ decided before $m$. As $m$ will be decided afterwards, its timestamp will be made greater than that of $m'$. Therefore, the barriers requested in order to deliver $m$ might not suffice anymore, requiring new, higher ones.

An approach to reduce this problem could be by using composite timestamps. Instead of relying on a single value -- apart from the message id for breaking ties --, two values could be used: one related to the instant when the message was created and one sequence number incremented only when the timestamp should be increased. The former we call $rtc$, standing for real-time clock, and the latter we call $seq$. Being $m$ any message, $m.ts = (rtc,seq)$. The second value is used only to break ties when the first one is equal between some two messages. In the case when the timestamp of $m$ has to be made greater than that of some message $m'$, where $m'.ts = (rtc', seq')$, the timestamp of $m$ would be changed to $(rtc', seq'+1)$.

In reply to the barrier request for some message $m : m.ts = (rtc,seq)$, a process in $blockers(m)$ would enqueue some message $b : b.ts = (rtc + 1, \text{---})$ to be sent to $m.src$. The whole purpose of this is: if two messages $m$ and $m'$, with initial timestamps $m.ts < m'.ts$ had barriers requested at the same time as their creation, even if $m.ts$ is changed because of $m'$, the barrier received for the delivery of $m'$ will probably also serve $m$ -- except when \mbox{$blockers(m) \setminus blockers(m') \neq \emptyset$} or when an empty message is enqueued, since no barrier is requested for empty messages, but they can still cause timestamp changes in other messages

@@

When employing the optimistic delivery proposed in Section \ref{sec:optdel}, some messages may still need to have their initial timestamp changed as a possible consequence of the optimistic assumption \emph{A2} not holding. Thus it may be necessary to request new, higher barrier values, resulting in longer delivery latencies, not only for the messages whose timestamp changed, but also for those after them in the delivery queue. Besides, delivering messages out of the initial timestamp order may increase the number of mistakes made by the optimistic delivery. A simple way to eliminate the possibility of having to change timestamps is by discarding messages that arrived too late, that is, with a timestamp lower than that of some already delivered message.

Although discarding messages which do not follow the initial timestamp order conflicts with the properties we have defined for the \amcast{} primitive, we describe this approach here because it brings the benefit of decreased number of mistakes and lower message delivery latency. The validity property would have to be changed to:

\begin{itemize}
  \item[(i)] If a correct process $p$ \amcast{}s $m$ and the optimistic assumption \emph{A2} holds, then  every correct process $q \in m.dst$ \amdel{}s $m$ \emph{(optimistic uniform validity)}.
\end{itemize}

%\textbf{Optimistic Uniform Validity}: If a process \amcast{}s $m$ and the optimistic assumption \emph{A2} holds, then one of the correct processes that is a destination of $m$ eventually \consdel{}s $m$.

The problem with changing the original uniform validity property is that it could make the multicast primitive impracticable for some applications. Nevertheless, some other applications not only do not require messages to be delivered if they arrive out of order, but also it is better not to deliver such messages at all -- such as real-time streams of audio or video or online real-time multiplayer games. In any case, the application could specify whether each message might be discarded or not.

\subsection{Complexity analysis}
\label{sec:complexity}

In this section, we analyse the time complexity of our atomic multicast protocol. We consider an optimized version of the algorithm described in Section \ref{sec:baseline}, that is, we are assuming that the optimizations suggested in Section \ref{sec:optimized} have been implemented. Therefore, our time complexity analysis considers that:

\begin{itemize}
  \item before proposing some message $m$, any process $p$ which proposes it waits for $w(p)$ time, after which $m$ can also be \optdel{}ed;
  \item consensus instances are run in parallel;
  \item consensus is implemented with Paxos and one of the processes of each group is elected as leader;
  \item when a process $p$ is \amcast{}ing some message $m$, before \rmcast{}ing it to the other processes of $group(p)$, $p$ sends barrier requests to all processes in $blockers(m)$;
  \item given any message $m$, when some process $p$ from $m.src$ know thats $m.ts$ has changed and that higher barrier values must be requested, $p$ sends such barrier requests.
\end{itemize}
